---
output:
  html_document: default
  pdf_document: default
---

---
title: "Homicides in the U.S.A."
output: html_document
---

Maria Poulli (24104745), James Paguay (23048419), Dio Koutras (23575500), Joann Lin (16006991)

```{r setup, include = FALSE}

# knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(modelr)
library(ggplot2)
library(rpart)
library(randomForest)
library(ISLR)
library(ModelMetrics)
library(gridExtra)
library(quantreg)

```

# Introduction

This project looks at homicides committed in the United States from 1980 onwards. Our analysis is centered around different age-groups and looks at certain sub-populations. We also look at the evolution of crime levels over time. Our random forest model tries to predict the race of perpetrators. Our linear model tries to predict the age of the perpetrator based on victim age. Our data was downloaded from Kaggle.

# About the Dataset

The dataset includes 638k rows and 24 columns. It contains extensive information about homicides such as location, type of crime, murder weapon, relationship between perpetrator and victim as well as age and race information for both victim and perpetrator. The columns that we decided to focus our work on are qualitative in nature with the exception of Age.

We reduced this dataframe to 420k rows and 16 columns by removing missing data in certain cells and by removing columns that contained information that was not pertinent to our analysis. We dropped the columns Agency_Code, Agency_Name and Agency_Type as they provided no information for our projections. The rows were reduced because of missing or incomplete data.

```{r, include = FALSE}

## Import database

database <- read_csv("database.csv")


## Check columns available
colnames(database)

## Only select relevant columns
homicides_columns <- database %>%
  select(State,
         Year,
         Month,
         Crime_Type,
         Crime_Solved,
         Victim_Sex,
         Victim_Age,
         Victim_Race,
         Perpetrator_Sex,
         Perpetrator_Age,
         Perpetrator_Race,
         Relationship,
         Weapon
         )

```

```{r clean_age, include = FALSE}

## Clean victim and perpetrator ages
homicides_clean <- homicides_columns %>%
  filter(Victim_Age != 998,
         !is.na(Perpetrator_Age),
         Perpetrator_Age > 4,
         Crime_Type != 'Manslaughter_by_Negligence') %>%
  arrange(Perpetrator_Age)

```

```{r clean_relationship, include = FALSE}

## Check values in Relationship
unique(database$Relationship)

## Clean Relationship column
## Change "Boyfriend" to "Boyfriend/Girlfriend"
homicides_clean$Relationship[homicides_clean$Relationship == "Boyfriend"] <- "Boyfriend/Girlfriend"

## Change "Girlfriend" to "Boyfriend/Girlfriend"
homicides_clean$Relationship[homicides_clean$Relationship == "Girlfriend"] <- "Boyfriend/Girlfriend"

## Change "Employee" to "Colleague"
homicides_clean$Relationship[homicides_clean$Relationship == "Employee"] <- "Colleague"

## Change "Employer" to "Colleague"
homicides_clean$Relationship[homicides_clean$Relationship == "Employer"] <- "Colleague"

## Change "Common-Law_Husband" to "Husband"
homicides_clean$Relationship[homicides_clean$Relationship == "Common-Law_Husband"] <- "Husband"

## Change "Common-Law_Wife" to "Wife"
homicides_clean$Relationship[homicides_clean$Relationship == "Common-Law_Wife"] <- "Wife"

unique(homicides_clean$Relationship)

```

```{r clean_weapon, include = FALSE}

## Check values in Weapon
unique(database$Weapon)

## Clean Weapon column
## Change "Gun" to "Firearm"
homicides_clean$Weapon[homicides_clean$Weapon == "Gun"] <- "Firearm"

## Change "Handgun" to "Firearm"
homicides_clean$Weapon[homicides_clean$Weapon == "Handgun"] <- "Firearm"

## Change "Rifle" to "Firearm"
homicides_clean$Weapon[homicides_clean$Weapon == "Rifle"] <- "Firearm"

## Change "Shotgun" to "Firearm"
homicides_clean$Weapon[homicides_clean$Weapon == "Shotgun"] <- "Firearm"

## Change "Suffocation" to "Suffocation/Strangulation"
homicides_clean$Weapon[homicides_clean$Weapon == "Suffocation"] <- "Suffocation/Strangulation"

## Change "Strangulation" to "Suffocation/Strangulation"
homicides_clean$Weapon[homicides_clean$Weapon == "Strangulation"] <- "Suffocation/Strangulation"

unique(homicides_clean$Weapon)

```

```{r clean_age_season, include = FALSE}

## Specify age groups for victims
categorize_victim_age <- function(x) {
  if(x <= 9) {
    return("9 and under")
  }
  else if (x <= 19){
    return("10-19")
  }
  else if (x <= 29) {
    return("20-29")
  }
  else if (x <= 39) {
    return("30-39")
  }
  else if (x <= 49) {
    return("40-49")
  }
  else if (x <= 59) {
    return("50-59")
  }
  else if (x <= 69) {
    return("60-69")
  }
  else if (x <= 79) {
    return("70-79")
  }
  else if (x <= 89) {
    return("80-89")
  }
  else if( x >= 90) {
    return("90 and over")
  }
}
homicides_clean$Victim_Age_Group=sapply(homicides_clean$Victim_Age, categorize_victim_age)

## Specify age groups for Perpetrator
categorize_perpetrator_age <- function(x) {
  if(x <= 9) {
    return("9 and under")
  }
  else if (x <= 19){
    return("10-19")
  }
  else if (x <= 29) {
    return("20-29")
  }
  else if (x <= 39) {
    return("30-39")
  }
  else if (x <= 49) {
    return("40-49")
  }
  else if (x <= 59) {
    return("50-59")
  }
  else if (x <= 69) {
    return("60-69")
  }
  else if (x <= 79) {
    return("70-79")
  }
  else if (x <= 89) {
    return("80-89")
  }
  else if( x >= 90) {
    return("90 and over")
  }
}
homicides_clean$Perpetrator_Age_Group=sapply(homicides_clean$Victim_Age, categorize_perpetrator_age)

## Categorise seasons
categorize_seasons <- function(x) {
  if(x=='December' | x=='January' | x=='February' ) {
    return("Winter")
  }
  else if (x=='March' | x=='April' | x=='May'){
    return("Spring")
  }
  else if (x=='June' | x=='July' | x=='August') {
    return("Summer")
  }
  else {
    return("Fall")
  }
}
homicides_clean$Season=sapply(homicides_clean$Month, categorize_seasons)

View(homicides_clean)

```
# Steps Take to Clean Data

Before proceeding with any analysis we cleand the Victim_Age, Perpetrator_Age, and Crime_Type columns. We did this by using the filter() function to retain all the rows within those columns that met our requirements. Thus, Victim ages equal to 998, Perpetrator ages that were blank or less than 4, and Crime type equal to Manslaughter by Negligence were filtered out.

Within the Relationship column, we reclassified "Boyfriend" and "Girlfriend" to "Boyfriend/Girlfriend", "Employee" and "Employer" to "Colleague", "Common-Law_Husband" to "Husband", and  "Common-Law_Wife" to "Wife". We did this by using the $ command to select the Relationship column within our dataset. Then, we applied subsetting by bracket notation to select the categories we wanted to change within Relationship. From there, we were able to apply the new categories to the selected categories by vectorization.

Next, we cleaned up the Weapon column with the same procedure we applied to the Relationship column. This allowed us to reclassify "Gun", "Handgun", "Rifle", and "Shotgun" to "Firearm", and "Suffocation" and "Strangulation" to "Suffocation/Strangulation".

Lastly, we created new columns of Perpetrator_Age_Group, Victim_Age_Group, and Seasons. To create a new column, we first defined a categorize function which includes a single multipath if statement. The if statement makes a logical comparison between our values from the selected column and what we expect by testing for a condition. It only returns a character string when the condition is True. Then, we used the sapply function with our selected column from our dataset and our categorize function as the arguments. We assigned the output from sapply to our new column, which was created by using the $ command on our dataset with the name our new column. This procedure was used to create our other two new columns.

# Time series of murders

Here are the first 5 years of homicides.

```{r exploratory_time_series_murders, include = FALSE}
time_series <- homicides_clean %>% 
  group_by(Year) %>%
  summarise(Murders_each_year = n()) %>%
  head(5)
time_series2 <- homicides_clean %>% 
  group_by(Year) %>%
  summarise(Murders_each_year = n())
```

```{r explanatory_time_series, echo = FALSE}
## Table - Time_Series
knitr::kable(
  time_series, 
  caption = "Evolution of murders over time",
  )

ggplot(data = time_series2) + geom_line(aes(x = Year,y = Murders_each_year))

```

## Overview of timeseries

We can see that there is a very sharp drop of murders committed starting in the early 1990's. The worst was at the beginning of the 80's and early 90's

**Incidents by State**

This graph showcases total incidents by state. As expected, the state with the highest population has the most murders. Further into our analysis, we used the data from California as it provided as solid foundation for our linear model.


```{r explanatory_incident_by_state, echo = FALSE}
ordered_state <- homicides_clean%>%
  group_by(State)%>%
  summarise(Total_murders = n(), .groups = 'drop')%>%
  arrange(desc(Total_murders))%>%
  ggplot() + geom_col(aes(x = reorder(State, - Total_murders), y = Total_murders))+
  labs(x= 'State', y = 'Incidents', title = 'Incidents by State')
ordered_state + theme(axis.text.x = element_text(angle = 90))
```

# Analysis of Victim vs Perp Sex

**Males vs Females**

We notice that Females are much more likely to be victims of murders. They are also less likely to be perpetrators.
```{r exploratory_victim_perp_sex, include = FALSE}
 vic_sex_proportion <- homicides_clean %>%
  select(Victim_Sex, Perpetrator_Sex) %>%
  group_by(Victim_Sex) %>%
  summarise(n_victime_sex = n()) %>%
  mutate(cum_total_victim = cumsum(n_victime_sex), pct = n_victime_sex/sum(n_victime_sex))

perp_sex_proportion<-homicides_clean %>%
  select(Victim_Sex, Perpetrator_Sex)%>%
  group_by(Perpetrator_Sex) %>%
  summarise(n_perp_sex = n()) %>%
  mutate(cum_total_perp = cumsum(n_perp_sex), pct = n_perp_sex/sum(n_perp_sex))


```

```{r overlay_perp_vic_sex, echo = FALSE}

ggplot(data = vic_sex_proportion, aes(Sex, proportion)) + geom_bar(aes(Victim_Sex,pct, color = 'vic_sex_proportion'), stat = 'identity', fill = 'grey', alpha = 0.5) +
    geom_bar(data = perp_sex_proportion, aes(Perpetrator_Sex, pct, color = 'perp_sex_proportion'), stat = 'identity',
             fill = "grey", alpha = 0.4)


```

# Analysis of Perpetrators by Age Groups

## Adults (18+) vs. Underage (under 18)

```{r exploratory_adults_vs_underage, include = FALSE}

## Total homicides
homicides_clean %>%
  summarize(n = n())
## 413155

## Adults
homicides_clean %>%
  filter(Perpetrator_Age >= 18) %>%
  summarize(Adults = n())
## 381302

## Underage
homicides_clean %>%
  filter(Perpetrator_Age < 18) %>%
  summarize(Underage = n())
## 31853

## Create tibble
perp_age_tibble <- tibble(
  Adults = 381302,
  Underage = 31853
)
perp_age_tibble

## Tidy data
perp_age_tidy <- perp_age_tibble %>%
  pivot_longer(c(`Adults`, `Underage`), names_to = "Perp_Age_Group", values_to = "N_homicides")
perp_age_tidy

## Add percentages
perp_age_percent <- perp_age_tidy %>%
  mutate(Percent_homicides = N_homicides / sum(N_homicides))
perp_age_percent

```

```{r exploratory_adults_vs_underage_show, echo = FALSE}

## Table
knitr::kable(
  perp_age_percent, 
  caption = "Adults vs. Underage",
  )

```

**Observations of Homicides Committed by Adults vs. Underage**

1. Adults perpetrators committed a total of 381302 homicides, which is about 92%.
2. Underage perpetrators committed a total of 31853 homicides, which is about 8%.

## Breakdown of Age Groups

```{r exploratory_breakdown_age_groups, include = FALSE}

## Breakdown by age groups
breakdown_age_groups <- homicides_clean %>%
  group_by(Perpetrator_Age_Group) %>%
  summarise(Count = n()) %>%
  arrange(desc(Count)) %>%
  mutate(Proportion = Count/sum(Count))

```

```{r exploratory_breakdown_age_groups_show, echo = FALSE}

## Table
knitr::kable(
  breakdown_age_groups, 
  caption = "Breakdown by Perpetrator Age Group",
  )

```

**Frequency of Homicides Committed within Age Groups**

1. The most frequent age group for perpetrators is 20-29, followed by 30-39 and then 10-19.
2. The least frequent age group for perpetrators is 9 and under followed by 90 and over.
3. From the age 40 to 90 the older you are, the least likely you are to commit crime.
4. 264 perpetrators belong in the "90 and over" Perpetrator_Age_Group. It has the second least total perpetrators within all groups. We will deep dive into interesting stats below.
5. 69972 belong to the group "10-19". It is the 3rd biggest age group in terms of perpetrators. Proportion is 17%. We will deep dive into interesting stats below.

## Elderly (90+) Perpetrators - Weapons of Choice

```{r exploratory_elderly_weapon, include = FALSE}

## Weapons
favorite_Weapon90Group <- homicides_clean %>%
  group_by(Perpetrator_Age_Group) %>%
  count(Weapon) %>%
  rename(Num_Weapons = n) %>%
  filter(Perpetrator_Age_Group == "90 and over") %>%
  mutate(Proportion = Num_Weapons / sum(Num_Weapons)) %>%
  arrange(desc(Num_Weapons))

## Weapons by sex
homicides_Perp_Race_Sex_Weapon <- homicides_clean %>%
  group_by(Perpetrator_Age_Group, Perpetrator_Race, Perpetrator_Sex, Victim_Age_Group, Weapon) %>%
  filter(Perpetrator_Age_Group == "90 and over") %>%
  summarise(count = n()) %>%
  arrange(desc(count))

## Weapons by sex overall
homicides_total_weapon_gender <-  homicides_clean %>%
  select(Weapon, Perpetrator_Sex) %>%
  group_by(Perpetrator_Sex, Weapon) %>%
  summarise(count = n()) %>%
  arrange(desc(count))

```

```{r exploratory_elderly_weapon_show, fig.show = "hold", out.width = "50%", echo = FALSE}

## Table - Weapons
knitr::kable(
  favorite_Weapon90Group, 
  caption = "Weapons of Choice by Elderly Perpetrators",
  )
## Table - Weapons overall


## Plot and comparison with overall dataset
ggplot(data = homicides_Perp_Race_Sex_Weapon) + 
  geom_col(mapping = aes(x = count, y = Weapon, fill = Perpetrator_Sex), position = "dodge") +
  labs(x = "Number of Crimes Committed", y = "Weapon Used", title = "90 and Over Perp. Group Favorite Weapons by Sex", fill = "Perpetrator Sex")

ggplot(data = homicides_total_weapon_gender) + geom_col(aes(count,Weapon, fill = Perpetrator_Sex), position = "dodge") +
  labs( x = 'Number of crimes committed', y = 'Weapon used', title = 'Weapon used by sex for all the dataset')


```

**Observations of Weapons of Choice**

1. The most common weapon used was a firearm followed by a blunt object. 179 people used Firearms followed by 33 people using a blunt object.
2. Males favored firearms while females favored blunt objects.

## Elderly (90+) Perpetrators - Victims Age

```{r exploratory_elderly_victim_age, include = FALSE}

## Victims age
victim_of_elderly <- homicides_clean %>%
  group_by(Perpetrator_Age_Group) %>%
  count(Victim_Age_Group) %>%
  rename(Num_VGroup = n) %>%
  filter(Perpetrator_Age_Group == "90 and over") %>%
  arrange(desc(Num_VGroup))

```

```{r exploratory_elderly_victim_age_show, echo = FALSE}

## Table - Victims age
knitr::kable(
  victim_of_elderly, 
  caption = "Victims Age of Elderly Perpetrators",
  )

```

**Observations of Victims Age**

1. The most targeted Victim_Age_Group was "80-89" with 76 victims within that group.
2. It was followed by "20-29" with 46 victims.

## Elderly (90+) Perpetrators - Perpetrators Sex

```{r exploratory_elderly_perp_sex, include = FALSE}

## Perpetrators sex
homicides_Perp_Sex <- homicides_clean %>%
  group_by(Perpetrator_Age_Group, Perpetrator_Sex, Victim_Age_Group, Weapon) %>%
  filter(Perpetrator_Age_Group == "90 and over") %>%
  summarise(Count = n()) %>%
  arrange(desc(Count))%>%
  head(7)

```

```{r exploratory_elderly_perp_sex_show, echo = FALSE}

## Table - Perpetrators sex
knitr::kable(
  homicides_Perp_Sex, 
  caption = "Sex of Elderly Perpetrators",
  )

```

**Observations of Perpetrators Sex**

1. Males killed more people between the ages of "80 - 89" using a firearm (56).
2. Females killed the most people between the ages of "90 and over" using a blunt object(2).
3. Females did not kill anyone in the Victim_Age_Group of "9 and under" as well.
4. Males killed at least 1 person in all Victim_Age_Group categories.

## Pre-Teen / Teen Perpetrators (10-19) - Teen Adults

```{r exploratory_teen_adults, include = FALSE}

teen_adults <- homicides_clean %>%
  filter(Perpetrator_Age==18 | Perpetrator_Age == 19) %>%
  count(Perpetrator_Age)

## Percentage
38257/69972

```

```{r exploratory_teen_adults_show, echo = FALSE}

## Table - Teen adults
knitr::kable(
  teen_adults, 
  caption = "Teens Adults",
  )

```

**Observations of Teens Adults**

Out of 69972, 38,257 are of age 18 or 19, i.e adults. Therefore 55% of the perpetrators in the group 10-19 are adults. 

## Pre-Teen / Teen Perpetrators (10-19) - Weapons of Choice

```{r exploratory_teen_weapon, include = FALSE}

## Weapons
favorite_Weapon10Group <- homicides_clean %>%
  group_by(Perpetrator_Age_Group) %>%
  count(Weapon) %>%
  rename(Num_Weapons = n) %>%
  filter(Perpetrator_Age_Group == "10-19") %>%
  mutate(Proportion = Num_Weapons / sum(Num_Weapons)) %>%
  arrange(desc(Num_Weapons)) %>%
  head(7)

```

```{r exploratory_teen_weapon _show, echo = FALSE}

## Table - Weapons
knitr::kable(
  favorite_Weapon10Group, 
  caption = "Weapons of Choice by Pre-Teen / Teen Perpetrators",
  )

```

**Observations of Weapons of Choice**

Most common weapon used was a firearm followed by a knife.  48661 used firearm, 10333 knife.


## Pre-Teen / Teen Perpetrators (10-19) - Victims Age

```{r exploratory_teen_victim_age, include = FALSE}

## Victims age
victim_of_teen <- homicides_clean %>%
  group_by(Perpetrator_Age_Group) %>%
  count(Victim_Age_Group) %>%
  rename(Num_VGroup = n) %>%
  filter(Perpetrator_Age_Group == "10-19") %>%
  arrange(desc(Num_VGroup))

```

```{r exploratory_teen_victim_age_show, echo = FALSE}

## Table - Victims age
knitr::kable(
  victim_of_teen, 
  caption = "Victims Age of Pre-Teen / Teen Perpetrators",
  )

```

**Observations of Victims Age**

1. The most targeted Victim_Age_Group was 20-29.
2. It was followed by 10-19.

## Pre-Teen / Teen Perpetrators (10-19) - Perpetrators Sex

```{r exploratory_teen_perp_sex, include = FALSE}

## Perpetrators sex
homicides_Perp_Sex_teen <-homicides_clean %>%
  group_by(Perpetrator_Age_Group, Perpetrator_Sex, Victim_Age_Group, Weapon) %>%
  filter(Perpetrator_Age_Group == "10-19") %>%
  summarise(Count = n()) %>%
  arrange(desc(Count)) %>%
  head(8)

```

```{r exploratory_teen_perp_sex_show, echo = FALSE}

## Table - Perpetrators sex
knitr::kable(
  homicides_Perp_Sex_teen, 
  caption = "Sex of Pre-Teen / Teen Perpetrators",
  )

```

**Observations of Perpetrators Sex**

1. Males killed more people between the ages of "20-29" using a firearm (16652).
2. Females killed more people between the ages of "10-19" using a firearm, followed by the ages "0-9" using a blunt object.

# Modeling

## Random Forest to Predict Perpetrator's Race

We created a Random Forest model to predict a perpetrator's race based on state, month, perpetrator sex, perpetrator age, weapon, victim sex, victim age, victim race, and relationship.  We hypothesize that the perpetrator's race is very closely related to the victim's race.

As a side note, we set the ntree = 100 for faster run time.

```{r random_forest_perpetrator_race, include = FALSE}

## Create indices.  Split homicides_clean dataset into 80% training data and 20% test data.
homicides_indices <- sample(1:nrow(homicides_clean), 0.8 * nrow(homicides_clean))
homicides_train <- homicides_clean %>% slice(homicides_indices)
homicides_test <- homicides_clean %>% slice(-homicides_indices)

## Fit random forest.
homicides_forest <- randomForest(as.factor(Perpetrator_Race) ~
                                   State +
                                   Month +
                                   Perpetrator_Sex +
                                   Perpetrator_Age +
                                   Weapon +
                                   Victim_Sex +
                                   Victim_Age +
                                   Victim_Race +
                                   Relationship,
                                 data = homicides_train,
                                 ntree = 100,
                                 importance = TRUE,
                                 do.trace = 10)


homicides_forest

## Identify the most important variables.
imp_hf <- importance(homicides_forest, type = 1)
tib_imp_hf <- imp_hf %>% as_tibble(rownames = "Variable") %>% arrange(desc(MeanDecreaseAccuracy))

```

```{r random_forest_important_variables_show, echo = FALSE}

## Table
knitr::kable(
  tib_imp_hf, 
  caption = "Important Variables",
  )

```

**Observations of Random Forest Model - Important Variables**

1. Victim_Race is the most important variable, followed by Relationship, Victim_Age, and Perpetrator_Age.
2. The Least important variable is Month.
3. Surprisingly, Victim_Sex is the second least important.

```{r random_forest_perpetrator_race_actual_vs_pred, include = FALSE}

## See actual Perpetrator_Race vs. predicted Perpetrator_Race.
results_hf <- homicides_test %>%
  add_predictions(homicides_forest, var = "pred_perp_race") %>%
  select(Perpetrator_Race,
         Victim_Race,
         Relationship,
         Victim_Age,
         Perpetrator_Age,
         Weapon,
         State,
         Perpetrator_Sex,
         Victim_Sex,
         Month,
         pred_perp_race)
results_hf %>% View()

## Calculate error rate
error_rate_results_hf <- results_hf %>% group_by(Perpetrator_Race) %>% 
  summarize(Count = n(), Wrong = sum(Perpetrator_Race != pred_perp_race), Error_rate = Wrong / Count) %>%
  mutate(Overall_error_rate = sum(Wrong)/sum(Count))

## Create confusion matrix.
pred_homicides_test <- homicides_test %>% add_predictions(homicides_forest, var = "pred_perp_race")
cm_hf <- xtabs(~ Perpetrator_Race + pred_perp_race, data = pred_homicides_test)
cm_hf

```

```{r random_forest_error_rate_confusion_matrix_show, echo = FALSE}

## Table
knitr::kable(
  error_rate_results_hf, 
  caption = "Error Details",
  )

## Table
knitr::kable(
  cm_hf, 
  caption = "Confusion Matrix",
  )

```

**Observations of Random Forest Model - Error Rate and Confusion Matrix**

1. The Error Rate of the model is approximately 0.13.
2. The model is most accurate when Perpetrator_Race = White or Black.
3. The model is least accurate when Perpetrator_Race = Asian/Pacific_Islander, Native_American/Alaska_Native, or Unknown.

## Linear Regression to Predict Victim's Age Based on Perpetrator's Age

Let's first look at the relative age (Victim - Perp) to see if there a pattern we can identify. We are using California for this analysis.

```{r explanatory_rel_age, include = FALSE}
filtered_homicides_rel <- homicides_clean %>%
  filter(State == "California" & Victim_Age > 9 & Perpetrator_Age > 9 & Victim_Age < 90 &
           Perpetrator_Age < 90 & Weapon == 'Firearm') %>%
  mutate(Relative_Age = Victim_Age - Perpetrator_Age)

```


```{r explanatory_boxplot_relative, echo = FALSE}

ggplot(data = filtered_homicides_rel)  +
  geom_boxplot(aes(x = Victim_Age, y = Relative_Age, fill = Perpetrator_Age_Group)) +
  labs (title = 'Relative Age Graph for Firearm Incidents, CA')
```

We notice that up to the age of 40, the victim and perpetrator are very similar in age. The median of each of those groups is almost equal to 0. The older they get the higher the relative age gets; this means that older victims tend to be targeted by individuals younger than they are. One possible interpretation would be that as the pool of potential victims gets older it is difficult to be targeted by someone even older than them, which would lead to a positive relative age. We also notice the spread from the boxplot graph that gets wider the further we move up the age range. The last age group is mostly outliers as data is sparse and does not follow the same pattern as the rest. 

**Linear Model**

We wanted to test our initial hypothesis that Victim_Age and Perp_Age follow a linear trend. We created a linear regression model to predict the perpetrator age based on victim age and other variables for the state of California.  Using the data from only one state allowed us to create clearer visuals.  We decided to focus on California because it had the most murder cases, so we would have a lot of data to work with. However, when we plot Perpetrator_Age by Victim_Age, we see that there are many outliers, specifically for age groups of "9 and under" and "90 and over" between Victim and Perpetrator groups. For ages of 99, this also represents ages greater than 99. So when plotted, we saw a strange pattern for that age.  Thus, we filtered out these age values to get a better understanding of the linear relation between Victim_Age and Perpetrator_Age. We also filtered out "9 and under" because there were very few data points in that age group, which skewed the results.

```{r linear_regression_cali_scatter_plot, include = FALSE}

filtered_homicides <- homicides_clean %>%
  filter(State == "California" & Victim_Age > 9 & Perpetrator_Age > 9 & Victim_Age < 90 &  Perpetrator_Age < 90)

```

```{r linear_regression_scatter_plot, echo = FALSE}

ggplot(filtered_homicides, aes(x = Victim_Age, y = Perpetrator_Age)) +
  geom_point(aes(color = Victim_Age_Group))

```

**Observations of Age Groups Data - Scatter Plot**

1. The Scatter Plot shows the Perpetrator Age by Victim Age relationship for each Victim Age Group.
2. The data looks scattered but you can see a small sign of linear relationship between our two variables. This relationship is not sufficiently clear because there are so many outliers within each Victim groups. To see a clearer linear trend would mean to remove the outliers.

```{r linear_regression_box_plot, echo = FALSE}

ggplot(filtered_homicides, aes(x = Victim_Age, y =  Perpetrator_Age, fill = Victim_Age_Group)) +
  geom_boxplot()

```

**Observations of Age Groups Data - Box Plot**

1. As Victim_Age increases, the median Perpetrator_Age increases. This suggest that Victim_Age is a great predictor variable for Perpetrator_Age, since there seems to be a linear regression. We also noticed how the older the Victim is, the wider the spread in Perpetrator_Age is. For example, the median Perpetrator_Age for Victim_Age_Group 20-29 is about 25 years old. This suggests that Perpetrators around 25 years old are more likely to commit a murder within their age range rather than outside it. However, for Victim_Age_Group 80-89, the median Perpetrator_Age is about 40. There is a wider spread of Perpetrator_Age, within this Victim_Age_Group.

```{r lm_perpetrator_age, include = FALSE}

homicides_indices <- sample(1:nrow(filtered_homicides), 0.8*nrow(filtered_homicides))
homicides_train <- filtered_homicides %>% slice(homicides_indices)
homicides_test  <- filtered_homicides %>% slice(-homicides_indices)

filtered_homicides_lm <- lm(Perpetrator_Age ~ Victim_Age * 
                              Victim_Race + 
                              Victim_Sex + 
                              Relationship + 
                              Weapon + 
                              0, data = homicides_train)

summary(filtered_homicides_lm)
## Shows the overall values mentioned above.

predict(filtered_homicides_lm)
## Gives us the predicted Victim_Age based on our linear model.

homicides_pred <- homicides_train %>% as_tibble() %>%
  mutate(pred_Page = predict(filtered_homicides_lm, newdata = homicides_train)) %>%
  select(Victim_Age_Group, Victim_Age, Perpetrator_Age, pred_Page)
## Adds the predicted ages to a tibble with the other selected columns.

homicides_pred

```

```{r linear_regression_perpetrator_age, echo = FALSE}

homicides_pred %>%
  ggplot(aes(x = Victim_Age, y = Perpetrator_Age, color = Victim_Age_Group )) + 
  geom_point() +
  geom_smooth(method = "lm", formula = "y ~ x", se = FALSE, span = 50)

```

**Observations of Linear Regression Model**

We started with Perpetrator_Age as the predictor variable. We then added other categorical predictor variables based on their p-value to increase our Adjusted R squared. Variables with a p-values greater than 0.05 were removed from our linear model because this meant there was greater probability that the variable will not be meaningful for our regression model. Since they did not improve our Adjusted R-squared, there was no need to keep those variables. In the end, our r-squared is about 0.89.

Since we removed the baseline and included the interaction in our linear model, we get different slopes for each Victim_Age_Group. As Perpetrator_Age increases, Victim_Age increases as well. However, we still see many of Victim_Age values outside of our slope of predicted values. This is because Homicide is motivated by a large number of factors making the victim somewhat random, even taking relationship, ages, etc. into account.
However, we still see between each Perpetrator_Age_Group there is still a linear pattern between ages of Perpetrators and Victims.

```{r linear_regression_residuals, echo = FALSE}

plot(filtered_homicides_lm$residuals)

```

**Observations of Linear Regression Model Residuals**

A good way to test the accuracy or fit of our model is to look at the residuals or the differences between the real values and the predicted values.  The idea here is that the sum of the residuals is approximately zero or close to it as possible.  Ideally, when you plot the residuals, they should look random. 
Otherwise, there might be a hidden pattern that the linear model is not considering.

When we look our residuals, we can see that majority of it is close to zero. However, we can see that some reach 20 and 60 in both directions. Residuals that are 60 could be outliers as there are not many of them. Residuals that fall between 20 and 40 in both directions are concerning, since there are a few. This could indicate that there is a hidden pattern our model cannot capture well. This was expected based on what we previously stated: Victims can be random regardless of their relationship, age, race, etc.

# Summary

Our analysis provided a lot of interesting facts and patterns about homicides in the United States. A positive observation is that since 1990 the murder rate has dropped sharply and significantly. As expected, most crimes are committed by adults and the most frequent age group for perpetrators is 20-29 and the most targeted Victim_Age_Group was 20-29. The 20-29 age group represents over a third of the dataset. Moreover, the 90&over do not seem to have a different preference compared to the general population in terms of weapon choice.

Turning to our two models, for the random forest, we observe that it offers a low error for the prediction of Perp_Race. For our linear model, we tried to predict the perpetrator age based on victim age and other variables for the state of California. Our regression has a high explanatory power and shows that there is a clear linear trend in the data. Our residuals graph also further strengthens our model by not exhibiting any patterns.

# Conclusion

Using different analysis methods we were able to extract and observe many different patterns and insights about the homicides in the United States. Our hypothesis was that perpetrators are usually the same race and similar age to that of the victim.  Given the analysis we conducted, we conclude that in an unsolved murder, the perpetrator is likely to be male, of the same race, and similar age group to the victim.

